Synthesizing Quotient Lenses

### 2018-03-12 papers/QuotientLenses/main.tex
Edited related work section

\begin{figure}[t]
\centering
\[
\begin{array}{l@{}l@{\ \ }l}
w \; \equiv_R \; w' &\iff& w = w' \\
w \; \equiv_{\project{R}{s}} \; w' && \text{for all }w, w'\\
w \; \equiv_{\squash{Q_1}{Q_2}{f}} \; w' &\iff& f(w) \equiv_{Q_2} w'
\text{ or } w \equiv_{Q_2} w' \\
w \; \equiv_{\normalize{R}{R'}{f}} \; w' &\iff&
f(w)=f(w') \text{ or }w = w'\\
w \; \equiv_{Q_1 \; ; \; Q_2} \; w' &\iff& \exists k, k' \in
\mathcal{L}(K(Q_2)) \text{ such that } w \; \equiv_{Q_1} \; k, \; w' \;
\equiv_{Q_1} \; k', \text{ and } k \; \equiv_{Q_2} \; k'\\
w \; \equiv_{Q_1 \cdot Q_2} \; w'  &\iff& w = r_1
\cdot r_2, \; w' = {r'}_1 \cdot {r'}_2 \text{ with } r_1 \; \equiv_{Q_1}
\; {r'}_1, \; r_2 \; \equiv{Q_n} \; {r'}_2\\
w \; \equiv_{Q_1 \sep Q_2} \; w' &\iff& w \; \equiv_{Q_1} \; w'
\text{ or } \; w \; \equiv_{Q_2} \; w'\\
w \; \equiv_{Q^*} \; w' &\iff& w = r_1 \cdot \ldots \cdot r_n, \; w'
= {r'}_1 \cdot \ldots \cdot {r'}_n \text{ and } r_i \equiv_{Q} \; {r'}_i
\\
w \; \equiv_{\perm{Q_1, \ldots, Q_n}{Q}} \; w' &\iff& w = r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}, \;
w' = {r'}_{\theta(1)} \cdot s'_1 \cdot \ldots \cdot s'_{n-1}
\cdot {r'}_{\theta(n)} \\
& & \text{for some } \sigma, \theta \in S_n, \text{ with } r_i \;
\equiv_{q_i} \; r'_i \text{ and } s_k \; \equiv_{Q} \; s'_{k}
\end{array}
\]
\caption{QRE Equivalence Relations}
\label{fig:relations}
\end{figure}
\subsection{Well Formed QREs}
\subsubsection{Completeness of {\em normalize}}
Before we present the full inference rules, for QREs, we discuss the {\em
normalize} QRE which is special in two main ways, the first of which has to do
with the premises in its inference rule:
\begin{prooftree}
\AxiomC{$\mathcal{L}(R') \subseteq \mathcal{L}(R)$}
\AxiomC{$f : \mathcal{L}(R) \longrightarrow \mathcal{L}(R')$}
\AxiomC{$f$ is surjective}
\AxiomC{$f = f^2$}
\RightLabel{(Normalize)}
\QuaternaryInfC{$\wf{\normalize{R}{R'}{f}}$}
\end{prooftree}
In order for the {\em normalize} QRE to be well-formed, then canonizing function
$f$ used by this combinator must be surjective and idempotent. Unfortunately,
verifying that a function satisfies these properties is in general undecidable.
Nonetheless, we allow a programmer to use normalize at their own risk without
checking these side conditions.

@@

Also notable is one of the data.gov examples, which converts demographic
statistics by zip code from xml form into rdf form. Synthesizing this
transformation certainly took the longest, nearly fifteen seconds, where the
next slowest took hardly over five. To validate this performance was not due to
our overhead, we modified these format to be bijective, so the synthesis task
could be input into Optician. Optician took over 15 seconds synthesize this
modified lens. We believe this conversion took longer because the input formats
were exceptionally large, requiring conversions between over 40 fields, not due
to an overhead from synthesizing from QREs.
\section{Related Work}
\label{relwork}

\iffalse
Bidirectional transformations have been used in diverse areas of computing
where they arise as parsers and pretty printers, marshallers
and unmarshallers, serializers and deserializers, database views and view
updaters and many others\sam{TODO put citations}. Such transformations have
been extensively studied since they were proposed as a solution to the classical
{\em view-update problem} in the database community, where the challenge is to
derive a program that extracts a view of data from a source, as well as a
program that folds view updates back into the source safely and correctly.
\fi
This paper builds on the work of Foster et al~\cite{quotientlenses} who
introduced the theory of quotient lenses and implemented quotient lenses as a
refinement of the bidirectional string processing language
Boomerang~\cite{boomerang}.  
In Boomerang, the source and target types are specified using regular
expressions, and the equivalence relations are expressed using canonizers, which
are functions that map elements of a regular language to their canonical
representative\bcp{If their use of the word is exactly what we've been using
throughout, then why define it?  Or, if we mean their exact set of
expressible canonizers, rather than the core concept of canonizer, can we be
precise about that distinction?}. Boomerang canonizers can express a very
broad class of 
equivalence relations between regular languages, but actually doing so is often
difficult with complicated equivalence relations. For instance, Boomerang's
in-build permutation combinator does not account for separators in between the
regular expressions\bcp{This sounds like a kind of trivial distinction.}. Also, this combinator permutes regular expressions rather
than canonizers, thus making it difficult to express nested permutations that
occur in many data formats, especially XML and XML-like formats.

Foster et al.\relax also discuss other bidirectional programming
languages that support quotienting of data including XSugar~\cite{xsugar},
biXid~\cite{bixid} and X/Inv~\cite{Hu2004,Mu2004,Mu2006}.\bcp{The fact that
  Fosater et al discuss these things is not germane to the present
  discussion.  We need to make our own comparisons (unless they are
  exactly the same, in which case we can just refer to Foster et al.).}
XSugar programs 
bidirectionally convert data stored in XML and ASCII formats respectively, with the
transformations specified by a pair of unambiguous grammars. The quotienting
occurs on the XML side by use of a generic canonizer that standardizes the
representation of trees. Well-formed XSugar programs are guaranteed to be
bijective modulo an equivalence relation that captures XML normalization. biXid
programs convert between pairs of XML documents, with the XML formats specified
using a pair of grammars as in XSugar. However, biXid grammars can be ambiguous.
This ambiguity is what allows biXid to express equivalences on the data.
Finally, Foster et al discuss a possible connection with the languages
X and Inv which support a primitive duplication combinator that does not work
well with the lens laws, but can be expressed using Boomerang quotient
lenses.\bcp{Same comment.}

\BCP{Are there really no bidirectional languages since 2008 that support any
form of quotienting??}

As far as synthesis goes, though there is a good deal of recent research on
synthesizing unidirectional string
transformations~\cite{singh2012learning,le-pldi-2014,gulwani-popl-2014,perelman2014test,Singh:blinkfill},
the system Optician, which we introduced in ~\cite{optician}, is the first attempt
to synthesize bidirectional transformations. In that
publication, we\bcp{confusing: Not the same ``we''!} compared Optician to
two of these unidirectional string 
transformers, Flash Fill~\cite{gulwani-popl-2014} and
FlashExtract~\cite{le-pldi-2014}, and found that these tools were unsuccessful
in synthesizing the complex transformatiunons that can be expressed in Optician.

\bcp{This paragraph seems redundant.}While Optician improved on prior effors
at synthesizing unidirectional string 
transformers and introduced synthesis for bidirectional string transformations,
Optician was only able to synthesize fully bijective data transformations. Our
new system \Name{} improves on Optician in that \Name{} loosens this
restriction, and is able to synthesize data which is bijective when put in
canonical form. \Name{} is therefore a refinement of Optician since the
new synthesis algorithm (i.e. the quotient lens synthesis algorithm) restricted
to ``pure'' regular expressions degenerates to the synthesis algorithm that
we introduced in Optician. Indeed, as we mentioned in the evaluation section,
\Name{} was able to synthesize all of Optician's benchmarks at a speed
competitive with Optician, though there was additional overhead in calculating
the whole language $W(c)$ and the kernel language $K(c)$ of QREs, which
resulted in a very slight decrease in performance.

Much of the research in synthesis assumes that the synthesizer is provided with
a collection of examples. Optician and Optometrist differ in that they requires
that the programmer supplies both examples {\em and} format descriptions in the
form of regular expressions or QREs.\bcp{But we are far from the only ones
  to consider type-based synthesis!!}  There is a trade-off here.  On the one
hand, a user must have some programming expertise to write regular expression
(or QRE) specifications, and it requires some work. On the other hand, such
specifications provide a great deal of information to the synthesis system,
which decreases the number of examples needed (often to zero), makes the system
scale well, and allows it to handle large, complex formats.  By providing these
format specifications, the synthesis engine does not have to both infer the
format of the data as well as the transformations on it, obviating the need to
infer tricky formats like those involving nested iterations. Furthermore, by
focusing on bidirectional transformations, we limit the space of synthesized
functions to bijective ones, reducing the search space, and the expressiveness
of the search space.

>>>>

\begin{figure}[t]
\centering
\[
\begin{array}{l@{}l@{\ \ }l}
w \; \equiv_R \; w' &\iff& w = w' \\
w \; \equiv_{\project{R}{s}} \; w' && \text{for all }w, w'\\
w \; \equiv_{\squash{Q_1}{Q_2}{f}} \; w' &\iff& f(w) \equiv_{Q_2} w'
\text{ or } w \equiv_{Q_2} w' \\
w \; \equiv_{\normalize{R}{R'}{f}} \; w' &\iff&
f(w)=f(w') \text{ or }w = w'\\
w \; \equiv_{Q_1 \; ; \; Q_2} \; w' &\iff& \exists k, k' \in
\mathcal{L}(K(Q_2)) \text{ such that } w \; \equiv_{Q_1} \; k, \; w' \;
\equiv_{Q_1} \; k', \text{ and } k \; \equiv_{Q_2} \; k'\\
w \; \equiv_{Q_1 \cdot Q_2} \; w'  &\iff& w = r_1
\cdot r_2, \; w' = {r'}_1 \cdot {r'}_2 \text{ with } r_1 \; \equiv_{Q_1}
\; {r'}_1, \; r_2 \; \equiv{Q_n} \; {r'}_2\\
w \; \equiv_{Q_1 \sep Q_2} \; w' &\iff& w \; \equiv_{Q_1} \; w'
\text{ or } \; w \; \equiv_{Q_2} \; w'\\
w \; \equiv_{Q^*} \; w' &\iff& w = r_1 \cdot \ldots \cdot r_n, \; w'
= {r'}_1 \cdot \ldots \cdot {r'}_n \text{ and } r_i \equiv_{Q} \; {r'}_i
\\
w \; \equiv_{\perm{Q_1, \ldots, Q_n}{Q}} \; w' &\iff& w = r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}, \;
w' = {r'}_{\theta(1)} \cdot s'_1 \cdot \ldots \cdot s'_{n-1}
\cdot {r'}_{\theta(n)} \\
& & \text{for some } \sigma, \theta \in S_n, \text{ with } r_i \;
\equiv_{q_i} \; r'_i \text{ and } s_k \; \equiv_{Q} \; s'_{k}
\end{array}
\]
\caption{QRE Equivalence Relations}
\label{fig:relations}
\end{figure}
\subsection{Well Formed QREs}
\label{subsec:well-formed-qres}
\subsubsection{Completeness of {\em normalize}}
Before we present the full inference rules, for QREs, we discuss the {\em
normalize} QRE which is special in two main ways, the first of which has to do
with the premises in its inference rule:
\begin{prooftree}
\AxiomC{$\mathcal{L}(R') \subseteq \mathcal{L}(R)$}
\AxiomC{$f : \mathcal{L}(R) \longrightarrow \mathcal{L}(R')$}
\AxiomC{$f$ is surjective}
\AxiomC{$f = f^2$}
\RightLabel{(Normalize)}
\QuaternaryInfC{$\wf{\normalize{R}{R'}{f}}$}
\end{prooftree}
In order for the {\em normalize} QRE to be well-formed, then canonizing function
$f$ used by this combinator must be surjective and idempotent. Unfortunately,
verifying that a function satisfies these properties is in general undecidable.
Nonetheless, we allow a programmer to use normalize at their own risk without
checking these side conditions.

@@

Also notable is one of the data.gov examples, which converts demographic
statistics by zip code from xml form into rdf form. Synthesizing this
transformation certainly took the longest, nearly fifteen seconds, where the
next slowest took hardly over five. To validate this performance was not due to
our overhead, we modified these format to be bijective, so the synthesis task
could be input into Optician. Optician took over 15 seconds synthesize this
modified lens. We believe this conversion took longer because the input formats
were exceptionally large, requiring conversions between over 40 fields, not due
to an overhead from synthesizing from QREs.
\section{Related Work}
\label{relwork}
This paper builds on the work of Foster et al~\cite{quotientlenses} who
introduced the theory of quotient lenses and implemented quotient lenses as a
refinement of the bidirectional string processing language
Boomerang~\cite{boomerang}. As we mentioned in
Section\cref{subsec:well-formed-qres}, all our QRE combinators can be expressed
using just the {\em normalize} combinator, which is one of the canonizer
primitives that Boomerang already supports. Also, all our QRE lens combinators
are already supported in Boomerang. Consequently Boomerang quotient lenses are
at least as expressive as our language of QRE lenses.

Boomerang's canonizers allow one to canonize a regular language $R$ to by
mapping it to another regular language $S$ which may not be contained in $S$.
Formally, given sets $C$ and $B$ and an equivalence relation on $B$, Foster et
al defined a {\em canonizer} $q$ from $C$ to $B/{\sim_B}$ to be a pair of
functions $q.\kw{canonize} : C \longrightarrow B$ and $q.\kw{choose} : B
\longrightarrow C$ such that for every $b \in B$,
$$q.\kw{canonize} \; (q.\kw{choose} \; b) \sim_B \; b$$
This definition gives allows much more latitude for defining canonizers than
QREs. For example, if $\sim_B = \kw{Tot}(B)$, the equivalence relation that
relates every element in $B$ to every other element in $B$, then every function
from $C$ to $B$ is a canonizer.

Because of this extra elbow-room, Boomerang has two primitive {\em duplication}
quotient lenses, the first of which can be derived using the following inference
rule,
\begin{prooftree}
\AxiomC{$\ell : C/{\sim_C} \Leftrightarrow A_1/{\sim_{A_1}}$}
\AxiomC{$f : C \longrightarrow A_2$}
\AxiomC{$A_1 \cdot^! A_2$}
\AxiomC{$\sim_A = \sim_{A_1} \cdot \kw{Tot}(A_2)$}
\QuaternaryInfC{$\kw{dup}_1 \ell \; f \; : \; C/{\sim_C} \Longrightarrow A_1
\cdot A_2/{\sim_A}$}
\end{prooftree}
\begin{align*}
(\kw{dup}_1 \; \ell \; f).\get \; c &= (\ell.\get \; c) \cdot (f \; c)\\
(\kw{dup}_1 \; \ell \; f).\lput \; (a_1 \cdot a_2) \; c &= \ell.\lput \; a_1 \; c\\
(\kw{dup}_1 \; \ell \; f).\create \; (a_1 \cdot a_2) &= \ell.\create \; a_1
\end{align*}
wiht the symmetric $\kw{dup}_2$ combinator discarding the first copy instead of
the second in the \lput/\create direction.

Boomerang's more general definition for canonizers also allows quotient lenses
to be used as canonizers by using the taking the \kw{canonize} function to be
the \get component of a lens and the \kw{choose} function to be its \create
component. Naturally, QREs also take advantage of this ability to use lenses as
canonizers by allowing for user-defined functions to be used by the \kw{squash}
and \kw{normalize} combinators. Moreover, the added functionality of
synthesizing lenses further makes it easier to define canonizers, as well as
lenses.

Foster et al also discuss other bidirectional programming languages that
support quotienting of data including XSugar~\cite{xsugar}, biXid~\cite{bixid}
and X/Inv~\cite{Hu2004,Mu2004,Mu2006}; we refer the reader to the related work
section in \cite{quotientlenses} for this discussion.

Another language which supports quotenting is BiFluX~\cite{pacheco2014biflux}, a
bidirectional functional update language for XML. BiFluX is inspired by
the FLUX XML update language~\cite{cheney2008flux}, and adopts a bidirectional
programming by update paradigm, where a program succintly and precisely
describes {\em how} to update a source document with a target document, in an
intuitive way, such that there is a unique ``inverse'' source query for each
update program. The source and view types of BiFluX programs are given by
regular expressions.

In BiFluX, a regular expression $R$ is said to be unambiguous if there is only
one way to parse a {\em flat} value of type $R$ to a stuctured value of type
$R$; here a flat value of $R$ is a value of $R$ with all left/right tags,
parantheses and list brackets removed. The bidirectional part of BiFluX involves
transforming source regular expressions to view regular expressions so that
data sequences described by the two types can be matched. This process can
result in intermediate types that are unambiguous, and this ambiguity can cause
unintended updates to be made the source, especially when information is
discared in this process. The typechecking phase of a BiFluX program therefore
includes a {\em type normalization} phase that tries to normalize a source types
using automata reduction techniques, and derive a lens between ambiguous and
unambiguous types. On the view side, this phase only tries to normalize view
types into isomorphic types.

{\em Formlenses}~\cite{rajkumar2014lenses} are a variant of lenses that perform
transformations up to equivalence. Formlenses are a bidirectional generalization
of {\em formlets}~\cite{cooper2008essence}. Formlets are a high-level
abstraction for building Web forms, and can be represented in Haskell as
functions that take a source of names---concretely, an \kw{Int}---as an
argument and produce a triple consisting of an HTML document, a {\em collect}
function, and a modified namesource:
\begin{center}
\begin{lstlisting}
type Formlet a = Int -> (Html, Env -> a, Int)
\end{lstlisting}
\end{center}
The names of any generated form fields are drawn from the namesource, and the
{\em collect} function looks up precisely these names.
Rajkumar et. al.~\cite{rajkumar2014lenses} observe that formlets are a useful
abstraction, but they only address one half of the problem---they make it easy
to construct a function that {\em produces} a value of type \kw{a}, but they do
not provide a way to describe a function that {\em consumes} a value of type
\kw{a} and embeds it in a form. A different way to think about this problem is
to observe that forms are often used to present an {\em updateable view} of the
data source. Therefore, Rajkumar et. al. leverage the power of bidirectional
transformations to define formlets that behave like updatable views, yielding
the abstraction of formlenses.

They \kw{Formless} type is given by,
\begin{center}
\begin{lstlisting}
type Formlens a = Maybe a -> [Int] -> (Html,Env -> Maybe a,[Int])
\end{lstlisting}
\end{center}
with the differences between this definition and the standard definition for
formlets being an extra \kw{Maybe a} parameter that provides an optional initial
value of the form, the namesource being changed to a list of integers, and the
{\em collect} function being allowed to return an optional value.

Rajkumar et. al. consider only formlenses that are well-formed in that a
well-formed formlens should only draw names from the namesource provided as an
argument, and the {\em collect} function should only look up corresponding
names.
Under this assumption, Rajkumar et. al. define what it means for a formlens to
be well behaved by defining a version of the classical GETPUT and PUTGET laws
for formlenses. These laws hold modulo an equivalence relation on environments,
since a formlens may be invoked with two separate namespaces, but this should
not affect the value stored in the information encapsulated in the type $a$.
Consequently, the GETPUT and PUTGET laws defined for formlenses are in spirit
analogous to the GETPUT and PUTGET laws for quotient lenses, since they hold up
to renaming of environment variables.

\iffalse
Well-behaved formlenses satisfy properties that are analogous to the classical
GETPUT and PUTGET law. If {\em extract} is a function from \kw{Html} to \kw{Env} that
extracts an environment containing the names and values of all the form fields
from an HTML document, then a formlens $f$ satisfies the Acceptability law if
for any $v$ of type $a$, $n$ of type \kw{Int} and environment $e$ of type
\kw{Env} if $f$ applied to \kw{Just v} and $n$ yields an HTML file $h$ and a
{\em collect} function $c$ where \kw{extract h} is contained in $e$, then
collect $e$
\fi

Finally, in Cunha~\cite{cunha2010relational} uses relational algebra to
encompass many of the existing approaches to bidirectional programming. For
example, using $\cdot$ to express relational composition, the GETPUT law states
that $\get \cdot \lput \subseteq \sim_V \cdot \pi_1$, where $\pi_1$ is the
projection onto the first factor. In addition to its generality, this
``point-free'' approach to defining bidirectional transformations has the
advantage of creating equational proofs of lens laws that proceed by folding
and unfolding combinator laws.

Now we turn to related work in synthesis. Much of the research in synthesis
assumes that the synthesizer is provided with a collection of examples. Optician
and Optometrist differ in that they require that the programmer supplies both
examples {\em and} format descriptions in the form of regular expressions or
QREs, though we are far from the only ones to consider type-based synthesis.
There is a trade-off here.  On the one hand, a user must have some programming
expertise to write regular expression (or QRE) specifications, and it requires
some work. On the other hand, such specifications provide a great deal of
information to the synthesis system, which decreases the number of examples
needed (often to zero), makes the system scale well, and allows it to handle
large, complex formats.  By providing these format specifications, the 
synthesis engine does not have to both infer the format of the data as well as
the transformations on it, obviating the need to infer tricky formats like
those involving nested iterations. Furthermore, by focusing on bidirectional
transformations, we limit the space of synthesized functions to bijective ones,
reducing the search space, and the expressiveness of the search space.
